Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-05-12 paper.qmd
Fixed the cross reference

The visualization of the selected layout in the $9\text{-}D$ shows edges between the clusters (@fig-mnist-tri-proj (b2)). This supports the presence of small separations between clusters. Additionally, the data points are not uniformly distributed across the clusters, resulting in dense areas (@fig-mnist-tri-proj (b2)). Furthermore, the clusters shows non-linear shapes (@fig-mnist-tri-proj (b3)).

@@

The digit 1 of the MNIST dataset consists of $7877$ grayscale images of handwritten digits [@lecun2010]. Before further analysis, PCA was used to preprocess the data, where the first $10$ principal components, explaining 83% of the total variation, were selected. The objective is to select a reasonable \gD{} layout, representing the non-linear structure of the digit 1 dataset in $10\text{-}D$ (@fig-mnist-tri-proj (a)).

@@

By visualizing the model generated for tSNE in $10\text{-}D$ helps to assess the \gD{} layout. As shown in @fig-mnist-tri-proj (a1), the model provides the evidence for the non-linear structure of the digit 1 data in $10\text{-}D$. The model shows some quirks. The model's twisted pattern provides evidence for the $10\text{-}D$ data structure, which is not observed by \gD{} layout (@fig-mnist-tri-proj (a2)).

@@

<!--add langevitour screenshots and youtube animation link-->
```{r}
#| label: fig-mnist-tri-proj
#| fig-pos: H
#| fig-height: 4
#| fig-width: 12
#| out-width: 100%
#| fig-cap: "\\gD{} wireframe layout shows that there is a big non-linear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster. The MNIST digit 1 data ($n =  7877$ and $p = 10$) has a non-linear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE ($a_1 = 0.041$, $b = 659/585 \\text{ } (30, \\text{ }48)$). Video of the langevitour animation is available at <>."

@@

```{r}
#| label: fig-clust-mnist
#| fig-pos: H
#| fig-height: 5
#| fig-width: 15
#| out-width: 100%
#| fig-cap: "The tSNE layout of the MNIST digit 1 data shows a big non-linear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster in \\gD{} (a). The brushing feature in the linked plots helps in visualizing the closeness of the small cluster to the big cluster. Videos of the linked plots are available at <>."

>>>>

The visualization of the selected layout in the $9\text{-}D$ shows edges between the clusters (@fig-model-pbmc-author-proj (b2)). This supports the presence of small separations between clusters. Additionally, the data points are not uniformly distributed across the clusters, resulting in dense areas (@fig-model-pbmc-author-proj (b2)). Furthermore, the clusters shows non-linear shapes (@fig-model-pbmc-author-proj (b3)).

@@

The digit 1 of the MNIST dataset consists of $7877$ grayscale images of handwritten digits [@lecun2010]. Before further analysis, PCA was used to preprocess the data, where the first $10$ principal components, explaining 83% of the total variation, were selected. The objective is to select a reasonable \gD{} layout, representing the non-linear structure of the digit 1 dataset in $10\text{-}D$ (@fig-clust-mnist (a)).

@@

By visualizing the model generated for tSNE in $10\text{-}D$ helps to assess the \gD{} layout. As shown in @fig-clust-mnist (a1), the model provides the evidence for the non-linear structure of the digit 1 data in $10\text{-}D$. The model shows some quirks. The model's twisted pattern provides evidence for the $10\text{-}D$ data structure, which is not observed by \gD{} layout (@fig-clust-mnist (a2)).

@@

<!--add langevitour screenshots and youtube animation link-->
```{r}
#| label: fig-mnist-tri-proj
#| fig-pos: H
#| fig-height: 4
#| fig-width: 12
#| out-width: 100%
#| fig-cap: "\\gD{} wireframe layout shows that there is a big non-linear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster. The MNIST digit 1 data ($n =  7877$ and $p = 10$) has a non-linear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE ($a_1 = 0.041$, $b = 659/585 \\text{ } (30, \\text{ }48)$). Video of the langevitour animation is available at <>."
#| eval: false

@@

```{r}
#| label: fig-clust-mnist
#| fig-pos: H
#| fig-height: 5
#| fig-width: 15
#| out-width: 100%
#| fig-cap: "The tSNE layout of the MNIST digit 1 data shows a big non-linear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster in \\gD{} (a). The MNIST digit 1 data ($n =  7877$ and $p = 10$) has a non-linear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE ($a_1 = 0.041$, $b = 659/585 \\text{ } (30, \\text{ }48)$). Video of the langevitour animation is available at <>. The brushing feature in the linked plots helps in visualizing the closeness of the small cluster to the big cluster. Videos of the linked plots are available at <>."
